Demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2025-08-28 Manuscript/ant_cactus_SUBMISSION2.tex
intro

\newpage{}
\section*{Introduction} \label{res:introduction}
%% introduce multispecies mutualisms and sequential mutualisms with definitions
\revise{Mutualisms, species interactions where all participants receive net benefits, are key components of ecological communities and can enhance the fitness and population growth rates of the partners involved.  \linelabel{res:27}
While traditionally framed as pairwise (+/+) interactions, mutualisms are often embedded within multispecies communities in which a focal species interacts with multiple partner species.
These broader, network-structured interactions are characteristic of generalized or diffuse mutualisms, in which mutualists interact with multiple partners rather than engaging in strict one‐to‐one pairings \citep{Aizen2014, Prior2015}.
Growing evidence suggests that pairwise interactions may be poor predictors of the net effects of multi-species mutualism \citep{Afkhami2014,Palmer2010,Bascompte2009,Dattilo2014}. 
A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Within a mutualist guild, partner species often differ in the amount or type of goods or services they provide, making partner identity an important source of contingency in mutualism \citep{Stanton2003}. 
Partner diversity can arise either from simultaneous interactions or sequential ones, where individual hosts experience turnover in partner identity over time \citep{Djieto-Lordon2005, Ness2006}. \linelabel{res:L22}
Understanding how partner diversity and turnover affect the overall fitness benefits of mutualism is an open question, especially in the context of environmental variability, where different partners may confer benefits unevenly across space or time  \citep{rogers2014, Thibaut2012, Frederickson2005, Palmer2010}. \linelabel{res:L39}}

%% Look at mechanisms which show how diversity impacts demography
There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \citep{Barrett2015,Ushio2020,Winfree2020}.
First, when there is a consistent ranking of best to worst mutualists, a more diverse sample of the partner community may be more likely to include the best partner \citep{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \citep{Batstone2018}.  
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
\revise{Second, even within a mutualist guild, the benefits conferred by alternative partner species can vary by method, efficiency, or timing \citep{Stachowicz2005,Bronstein2006,Stanton2003}. \linelabel{res:L53}
This can lead to a positive effect of partner diversity through complementarity, whereby diversity increases the net benefits received by focal mutualist beyond what a single partner species could offer \citep{Batstone2018}.
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than expected from the sum of complementary functions \citep{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \citep{Batstone2018,Lazaro2022,Horvitz1990}. 
In the context of temporally fluctuating environments, the portfolio effect implies that inter-annual variation in partner identity could stabilize overall plant fitness across years. \linelabel{res:portfolio_asynchronous} 
For such a portfolio effect to operate, two conditions must be met: (i) partner species must differ in the timing or magnitude of their responses to environmental variation (i.e., their benefits fluctuate asynchronously), and (ii) the combined effect of these fluctuations must result in increased net benefit to the host across years relative to the net benefit expected if all partners responded in parallel \citep{Loreau1999, Tilman1994, Hautier2014}. \linelabel{res:portfolio_criteria}
While portfolio effects have been well studied in plant and animal communities, their role in mutualisms—particularly where partner diversity is expressed sequentially rather than simultaneously—remains less explored \citep{Ushio2020}.}

@@

%% Defensive ant-plant mutualisms
Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014} and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
\revise{Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}, however little is known about the combined effects of partner identity, directional partner turnover and temporal stochasticity, particularly because long-term demographic studies on ant-plant interactions are exceedingly rare \cite{Donald2022,Moraes2009}. }

\revise{Long-term demographic data are essential to understanding the fitness consequences of multi-species mutualisms for several reasons. \linelabel{res:long-term-data}
First, fitness effects of sequential partnering play out over long time scales due to the accumulation of partners across time rather than the effects of multiple partners at once. 
While turnover of mutualist partners is known to occur \citep{Ness2006, Djieto-Lordon2005,Fonseca2003}, long-term data documenting patterns and consequences of partner switching are uncommon, especially in field settings \citep{Palmer2010}. 
Second, certain effects of partner diversity are fluctuation-dependent (i.e., portfolio effect) and would only be detectable in the context of temporally stochastic environments in which alternative partners may fluctuate asynchronously. 
Quantifying contributions from the portfolio effect therefore requires long-term data on partner (a)synchrony across inter-annual fluctuations. 
In practical terms, evidence for a portfolio effect would include: 
(i) partner species showing asynchronous variation in their benefits to the focal host across years, and 
(ii) the combined effect of these fluctuations must increase the net benefits of partner diversity to the host relative to the variance expected if all partners responded in parallel. \linelabel{res:portfolio}
Detecting these patterns typically involves either correlation analyses of inter-annual partner benefits, or null-model simulations comparing observed host fitness variance to expectations under random or synchronous partner turnover.
Finally, alternative partners may have different demographic effects (e.g., one that enhances survival and another that enhances reproduction), and these effects may contribute differently to overall fitness of a focal mutualist.  
Most studies evaluate short-term fitness proxies, such as seed set or herbivory, often focusing on a single life stage \citep{Palmer2010, Trager2010}, while fewer trace demographic effects across multiple stages \citep{Bruna2014}.
Integrating long-term data from across the life cycle within a demographic modeling framework therefore provides a powerful opportunity to diagnose the fitness consequences of partner diversity and turnover in the context of environmental stochasticity.  
}

@@

%% Models and questions
Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover, and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
The long-term nature of this observational dataset gives us the power to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. \linelabel{res:long-term-data2}
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
\revise{We then built a stochastic, multi-state integral projection model (IPM) which combines diverse effects on vital rates and pathways of partner turnover. 
Finally, we used the model to conduct simulation `experiments' that quantify plant fitness under different scenarios of partner richness, composition, and synchrony.}

>>>>

\newpage{}
\section*{Introduction} \label{res:introduction}
%% introduce multispecies mutualisms and sequential mutualisms with definitions
\revise{Mutualisms, species interactions where all participants receive net benefits, are key components of ecological communities and can enhance the fitness and population growth rates of the partners involved.  \linelabel{res:27}
While traditionally framed as pairwise (+/+) interactions, mutualisms are often embedded within multispecies communities in which a focal mutualist interacts with multiple partner species.
%These broader, network-structured interactions are characteristic of generalized or diffuse mutualisms, in which mutualists interact with multiple partners rather than engaging in strict one‐to‐one pairings \citep{Aizen2014, Prior2015}. %tom: I would avoid discussion of 'networks'
Growing evidence suggests that pairwise interactions may be poor predictors of the net effects of multi-species mutualism \citep{Afkhami2014,Palmer2010,Bascompte2009,Dattilo2014}.

A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Within a mutualist guild, partner species often differ in the amount or type of goods or services they provide, making partner identity an important source of contingency in mutualism \citep{Stanton2003}. 
Partner diversity can arise either from simultaneous interactions or sequential ones, where individual hosts experience turnover in partner identity over time \citep{Djieto-Lordon2005, Ness2006}. \linelabel{res:L22}
Understanding how partner diversity and turnover affect the overall fitness benefits of mutualism is an open question, especially in the context of environmental variability, where different partners may confer benefits unevenly across space or time  \citep{rogers2014, Thibaut2012, Frederickson2005, Palmer2010}. \linelabel{res:L39}}

%% Look at mechanisms which show how diversity impacts demography
There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \citep{Barrett2015,Ushio2020,Winfree2020}.
First, when there is a consistent ranking of best to worst mutualists, a more diverse sample of the partner community may be more likely to include the best partner \citep{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \citep{Batstone2018}.  
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
\revise{Second, even within a mutualist guild, the benefits conferred by alternative partner species can vary by method, efficiency, or timing \citep{Stachowicz2005,Bronstein2006,Stanton2003}. \linelabel{res:L53}
This can lead to a positive effect of partner diversity through complementarity, whereby diversity increases the net benefits received by focal mutualist beyond what a single partner species could offer \citep{Batstone2018}.
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than expected from the sum of complementary functions \citep{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a `portfolio' that stabilizes fitness benefits across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \citep{Batstone2018,Lazaro2022,Horvitz1990}. 
In the context of temporally fluctuating environments, the portfolio effect implies that inter-annual variation in partner identity could stabilize overall plant fitness across years. \linelabel{res:portfolio_asynchronous} 
For such a portfolio effect to operate, two conditions must be met: (i) partner species must differ in the timing or magnitude of their responses to environmental variation (i.e., their benefits fluctuate asynchronously), and (ii) the combined effect of these fluctuations must result in increased net benefit to the host across years relative to the net benefit expected if all partners responded in parallel \citep{Loreau1999, Tilman1994, Hautier2014}. \linelabel{res:portfolio_criteria}
While portfolio effects have been well studied in plant and animal communities, their role in mutualisms—particularly where partner diversity is expressed sequentially rather than simultaneously—remains less explored \citep{Ushio2020}.}

@@

%% Defensive ant-plant mutualisms
Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014} and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
\revise{Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}, however little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because long-term demographic studies on ant-plant interactions are exceedingly rare \cite{Donald2022,Moraes2009}. }

\revise{Long-term demographic data are essential to understanding the fitness consequences of multi-species mutualisms for several reasons. \linelabel{res:long-term-data}
First, fitness effects of sequential partnering play out over long time scales due to the accumulation of partners across time rather than the effects of multiple partners at once. 
While turnover of mutualist partners is known to occur \citep{Ness2006, Djieto-Lordon2005,Fonseca2003}, long-term data documenting patterns and consequences of partner switching are uncommon, especially in field settings \citep{Palmer2010}. 
Second, certain effects of partner diversity are fluctuation-dependent (i.e., portfolio effect) and would only be detectable in the context of temporally stochastic environments in which alternative partners may fluctuate asynchronously. 
Quantifying contributions from the portfolio effect therefore requires long-term data on partner (a)synchrony across inter-annual fluctuations. 
%In practical terms, evidence for a portfolio effect would include: 
%(i) partner species showing asynchronous variation in their benefits to the focal host across years, and 
%(ii) the combined effect of these fluctuations must increase the net benefits of partner diversity to the host relative to the variance expected if all partners responded in parallel. \linelabel{res:portfolio}
%Detecting these patterns typically involves either correlation analyses of inter-annual partner benefits, or null-model simulations comparing observed host fitness variance to expectations under random or synchronous partner turnover.
Finally, alternative partners may have different demographic effects (e.g., one that enhances survival and another that enhances reproduction), and these effects may contribute differently to overall fitness of a focal mutualist.  
Most studies evaluate short-term fitness proxies, such as seed set or herbivory, often focusing on a single life stage \citep{Palmer2010, Trager2010}, while fewer trace demographic effects across multiple stages \citep{Bruna2014}.
Integrating long-term data from across the life cycle within a demographic modeling framework therefore provides a powerful opportunity to diagnose the fitness consequences of partner diversity and turnover in the context of environmental stochasticity.  
}

@@

%% Models and questions
Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover, and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
The long-term nature of this observational dataset gives us the power to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. \linelabel{res:long-term-data2}
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
\revise{We then built a stochastic, multi-state integral projection model (IPM) which combines diverse effects on vital rates and pathways of partner turnover. 
Finally, we used the model to quantify plant fitness under different scenarios of partner richness, composition, and synchrony.}
